---
title: "Data Processing"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Adult Income Dataset

In this chapter, we will explore and process the Adult Income dataset, which will serve as the foundation for our comparison of LIME and SHAP interpretation methods. This dataset is widely used in machine learning research and provides a practical benchmark for classification tasks.

## Dataset Overview

The Adult Income dataset (also known as "Census Income" dataset) was extracted from the 1994 U.S. Census Bureau database. The prediction task is to determine whether a person earns more than \$50,000 a year based on census data.

```{r load-libraries}
# Loading necessary libraries
library(tidyverse)  # For data manipulation and visualization
library(caret)      # For machine learning workflows
library(scales)     # For better visualization scales
library(DataExplorer) # For automated EDA
library(skimr)      # For data summaries
```

```{r load-data}
# Loading the Adult dataset
# We'll use the UCI Machine Learning Repository version
data_url <- "https://archive.ics.uci.edu/ml/machine-learning-databases/adult/adult.data"
column_names <- c("age", "workclass", "fnlwgt", "education", 
                 "education_num", "marital_status", "occupation", 
                 "relationship", "race", "sex", "capital_gain", 
                 "capital_loss", "hours_per_week", "native_country", "income")

adult <- read.csv(data_url, header = FALSE, strip.white = TRUE, 
                 col.names = column_names, na.strings = c("?", "", "NA", "N/A"))

# Display the first few rows
head(adult)
```

## Dataset Characteristics

The Adult Income dataset includes the following characteristics:

```{r dataset-info}
# Dataset dimensions
cat("Dataset dimensions:", dim(adult)[1], "rows and", dim(adult)[2], "columns\n")

# Class distribution
income_distribution <- table(adult$income)
cat("Class distribution:\n")
print(income_distribution)
cat("Percentage of high income (>50K):", 
    percent(income_distribution[2] / sum(income_distribution)), "\n")

# Quick summary of the dataset
skim(adult)
```

The dataset contains approximately 32,000 instances with 14 features (predictors) and 1 target variable (income). The class distribution shows an imbalance, with a higher proportion of individuals earning less than \$50,000 per year.

## Feature Analysis

Let's examine the features in more detail:

```{r feature-analysis}
# Categorical features
cat_features <- names(adult)[sapply(adult, is.factor) | sapply(adult, is.character)]
if (length(cat_features) == 0) {
  # Convert character columns to factors if none are already factors
  for (col in names(adult)[sapply(adult, is.character)]) {
    adult[[col]] <- as.factor(adult[[col]])
  }
  cat_features <- names(adult)[sapply(adult, is.factor)]
}

# Numeric features
num_features <- names(adult)[sapply(adult, is.numeric)]

# Summary of categorical features
cat("Categorical features:", length(cat_features), "\n")
print(cat_features)

# Summary of numerical features
cat("\nNumerical features:", length(num_features), "\n")
print(num_features)

# Missing values check
missing_values <- colSums(is.na(adult))
if (sum(missing_values) > 0) {
  cat("\nColumns with missing values:\n")
  print(missing_values[missing_values > 0])
} else {
  cat("\nNo missing values found.\n")
}
```

## Data Preprocessing

We'll now preprocess the data to make it suitable for our machine learning models:

```{r data-preprocessing}
# Create a copy of the dataset for preprocessing
adult_processed <- adult

# Handle missing values
if (sum(is.na(adult)) > 0) {
  # For categorical features, replace NA with the most frequent category
  for (col in cat_features) {
    if (sum(is.na(adult_processed[[col]])) > 0) {
      most_frequent <- names(sort(table(adult_processed[[col]]), decreasing = TRUE))[1]
      adult_processed[[col]][is.na(adult_processed[[col]])] <- most_frequent
      cat("Replaced NAs in", col, "with", most_frequent, "\n")
    }
  }
  
  # For numeric features, replace NA with the median
  for (col in num_features) {
    if (sum(is.na(adult_processed[[col]])) > 0) {
      median_val <- median(adult_processed[[col]], na.rm = TRUE)
      adult_processed[[col]][is.na(adult_processed[[col]])] <- median_val
      cat("Replaced NAs in", col, "with median:", median_val, "\n")
    }
  }
}

# Convert the target variable to a factor if it's not already
adult_processed$income <- factor(adult_processed$income, 
                               levels = c("<=50K", ">50K"), 
                               labels = c("low", "high"))

# Encode categorical variables
# We'll use one-hot encoding for our models
dummy_vars <- dummyVars(~ ., data = select(adult_processed, all_of(cat_features)), fullRank = TRUE)
categorical_encoded <- predict(dummy_vars, select(adult_processed, all_of(cat_features)))
saveRDS(dummy_vars, file = "data/dummy_encoder.rds")

# Combine with numeric variables
adult_encoded <- cbind(select(adult_processed, all_of(num_features)), 
                     as.data.frame(categorical_encoded),
                     income = adult_processed$income)

# Check the final preprocessed dataset
cat("Preprocessed dataset dimensions:", dim(adult_encoded)[1], "rows and", 
    dim(adult_encoded)[2], "columns\n")
```

## Feature Importance Visualization

Before we build our models, let's visualize the relationship between features and the target variable:

```{r feature-importance}
# For numeric features
if (length(num_features) > 0) {
  adult_long <- adult_processed %>%
    select(all_of(c(num_features, "income"))) %>%
    pivot_longer(cols = all_of(num_features), 
                names_to = "feature", 
                values_to = "value")
  
  ggplot(adult_long, aes(x = value, fill = income)) +
    geom_density(alpha = 0.5) +
    facet_wrap(~ feature, scales = "free") +
    theme_minimal() +
    labs(title = "Distribution of Numeric Features by Income Level",
         x = "Value", y = "Density") +
    scale_fill_brewer(palette = "Set1")
}

# For categorical features (selecting a few important ones to avoid overcrowding)
important_cat_features <- c("education", "marital_status", "occupation", "relationship")
important_cat_features <- intersect(important_cat_features, cat_features)

if (length(important_cat_features) > 0) {
  for (feature in important_cat_features) {
    p <- ggplot(adult_processed, aes_string(x = feature, fill = "income")) +
      geom_bar(position = "fill") +
      theme_minimal() +
      theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
      labs(title = paste("Income Distribution by", feature),
           y = "Proportion", x = feature) +
      scale_fill_brewer(palette = "Set1")
    print(p)
  }
}
```

## Strict Train-Test Split with CSV Export

To ensure there's no data leakage, we'll implement a strict train-test split strategy and save the datasets as separate CSV files:

```{r train-test-split}
# Set seed for reproducibility
random_seed <- 9876
set.seed(random_seed)
cat("Random seed used:", random_seed, "\n")

# Perform stratified sampling to maintain class distribution
train_indices <- createDataPartition(adult_encoded$income, p = 0.7, list = FALSE, 
                                     times = 1)  # Only create one split

# Create training and testing sets
train_data <- adult_encoded[train_indices, ]
test_data <- adult_encoded[-train_indices, ]

# Verify there's no overlap between training and testing sets
cat("Number of overlapping rows between train and test sets:", 
    sum(rownames(train_data) %in% rownames(test_data)), 
    "(should be 0)\n")

# Check class distribution in train and test sets
cat("Training set class distribution:\n")
print(table(train_data$income))
cat("Training set class percentages:\n")
print(prop.table(table(train_data$income)))

cat("\nTest set class distribution:\n")
print(table(test_data$income))
cat("Test set class percentages:\n")
print(prop.table(table(test_data$income)))

# Verify dimensions of splits
cat("\nTraining set dimensions:", dim(train_data)[1], "rows and", dim(train_data)[2], "columns\n")
cat("Testing set dimensions:", dim(test_data)[1], "rows and", dim(test_data)[2], "columns\n")

# Create data directory if it doesn't exist
if (!dir.exists("data")) {
  dir.create("data")
}

# Save training and test sets as CSV files
write.csv(train_data, file = "data/adult_train.csv", row.names = FALSE)
write.csv(test_data, file = "data/adult_test.csv", row.names = FALSE)

# Save random seed information for reference
seed_info <- data.frame(seed = random_seed)
write.csv(seed_info, file = "data/random_seed.csv", row.names = FALSE)

cat("Training and test sets have been saved as separate CSV files.\n")
cat("Random seed information has been saved as a CSV file.\n")
```

## Class Imbalance

The Adult Income dataset exhibits class imbalance, with significantly fewer instances of high income (\>\$50K) compared to low income (≤\$50K). This imbalance can bias machine learning models toward the majority class. Let's visualize this imbalance and create a balanced training set:

```{r class-imbalance}
# Visualize class imbalance
ggplot(adult_processed, aes(x = income, fill = income)) +
  geom_bar() +
  theme_minimal() +
  labs(title = "Class Distribution in Adult Income Dataset",
       x = "Income Level", y = "Count") +
  scale_fill_brewer(palette = "Set1") +
  geom_text(stat = 'count', aes(label = paste0(round(..count../nrow(adult_processed)*100, 1), "%")), 
            vjust = -0.5)
```

### Creating a Balanced Training Set

For better model performance, especially with neural networks, we'll create a balanced training dataset:

```{r balanced-dataset}
# Calculate class weights
class_weights <- table(train_data$income)
class_weights <- 1 / (class_weights / sum(class_weights))
class_weights <- class_weights / sum(class_weights) * 2
names(class_weights) <- levels(train_data$income)
print(class_weights)

# Create balanced sample
set.seed(random_seed)  # Use the same random seed
low_indices <- which(train_data$income == "low")
high_indices <- which(train_data$income == "high")

# Undersample the majority class
n_minority <- length(high_indices)
sampled_low_indices <- sample(low_indices, n_minority)

# Create balanced dataset
balanced_indices <- c(sampled_low_indices, high_indices)
balanced_train_data <- train_data[balanced_indices, ]

# Verify class distribution in balanced dataset
cat("Balanced training set class distribution:\n")
print(table(balanced_train_data$income))

# Save balanced training set as CSV
write.csv(balanced_train_data, file = "data/adult_train_balanced.csv", row.names = FALSE)
cat("Balanced training set has been saved as a CSV file.\n")
```

## Data Validation

To ensure our data processing pipeline is robust, we'll perform some additional validation checks:

```{r data-validation}
# Check if the two classes in the balanced training set are equal
balanced_class_counts <- table(balanced_train_data$income)
cat("Class counts in balanced training set:", balanced_class_counts[1], "vs", balanced_class_counts[2], "\n")
cat("Class ratio (low:high):", balanced_class_counts[1]/balanced_class_counts[2], "\n")

# Verify feature column name consistency across all datasets
cat("\nFeature column name consistency check:\n")
cat("Training set columns:", ncol(train_data), "\n")
cat("Testing set columns:", ncol(test_data), "\n")
cat("Balanced training set columns:", ncol(balanced_train_data), "\n")

# Check for NA values in each dataset
cat("\nNA values check:\n")
cat("Total NAs in training set:", sum(is.na(train_data)), "\n")
cat("Total NAs in test set:", sum(is.na(test_data)), "\n")
cat("Total NAs in balanced training set:", sum(is.na(balanced_train_data)), "\n")

# Check feature value ranges
cat("\nNumeric feature range check (train vs test):\n")
for(feat in num_features) {
  if(feat %in% colnames(train_data) && feat %in% colnames(test_data)) {
    cat(feat, "- Train range:", min(train_data[[feat]]), "to", max(train_data[[feat]]),
        "| Test range:", min(test_data[[feat]]), "to", max(test_data[[feat]]), "\n")
  }
}
```

## Summary

In this chapter, we've:

1.  Loaded and explored the Adult Income dataset
2.  Identified and analyzed its key features
3.  Preprocessed the data by handling missing values and encoding categorical variables
4.  Visualized important relationships between features and income levels
5.  Implemented a strict train-test split strategy with CSV exports to prevent data leakage
6.  Created a balanced training dataset to address class imbalance concerns
7.  Performed additional data validation checks to ensure robust data processing

The preprocessed data is now saved as CSV files, ready for model building and the application of interpretability methods, which we'll explore in the next chapter. This approach ensures clear separation between training and testing data, helping to prevent data leakage issues.
