---
title: "Enhanced Visualizations"
---

# Enhanced Visualizations for Interpretability

This chapter adds visually rich and engaging plots to help communicate model behavior and interpretability insights.

## 1. Correlation Heatmap

Visualizing correlations between numeric features in the dataset.

```{r}
library(corrplot)
library(dplyr)
train_data <- read.csv("data/adult_train.csv")
test_data <- read.csv("data/adult_test.csv")
rf_importance_df <- readRDS("models/rf_importance_df.rds")

top_vars <- rf_importance_df$Feature[1:5]

selected_data <- train_data[, top_vars]

selected_data <- selected_data %>% select(where(is.numeric))

cor_matrix <- cor(selected_data, use = "complete.obs")

corrplot(
  cor_matrix, method = "color", type = "upper",
  tl.col = "black", addCoef.col = "black",
  number.cex = 0.8, title = "Correlation Heatmap of Top Variables",
  mar = c(0, 0, 2, 0)
)


```

## 2. Parallel Coordinate Plot of Important Features

```{r, message=FALSE, warning=FALSE}
library(GGally)
top_rf_features <- rf_importance_df$Feature[1:6]  # top 6 for visibility

parallel_df <- train_data[, c(top_rf_features, "income")]

GGally::ggparcoord(data = parallel_df,
                   columns = 1:6,
                   groupColumn = 7,
                   scale = "std",
                   showPoints = TRUE,
                   alphaLines = 0.3) +
  theme_minimal() +
  labs(title = "Parallel Coordinate Plot of Top RF Features")
```

## 3. Dimensionality Reduction (UMAP)

```{r umap-plot, message=FALSE, warning=FALSE}
library(umap)
library(RColorBrewer)

umap_model <- umap::umap(select(train_data, -income))
umap_df <- as.data.frame(umap_model$layout)
umap_df$income <- train_data$income

# Plot
ggplot(umap_df, aes(x = V1, y = V2, color = income)) +
  geom_point(alpha = 0.6) +
  scale_color_brewer(palette = "Set1") +
  labs(title = "UMAP Projection of Training Data", x = "UMAP1", y = "UMAP2") +
  theme_minimal()
```

------------------------------------------------------------------------

These visualizations provide interactive and exploratory perspectives that can complement formal interpretability metrics. They are especially useful when presenting to non-technical audiences or during exploratory analysis.
